---
title: "Basics of the SpatialExperiment object"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basics of the SpatialExperiment object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SPIAT3D)
```

In this vignette, I want to shortly go through some of the specifics of using a
SpatialExperiment object. This includes how to make one, and the key features
of one that SPIAT3D exploits.

# Making a SpatialExperiment object
The following code is an example of how to convert a data frame into a
SpatialExperiment object. What if I don't have a data frame? Convert your data
into a simple data frame with the required data, then go from here.

For SPIAT3D, the only data required is the x,y,z coordinates of each cell type
and the cell type identity of each cell type (e.g. Tumour, Immune, etc.). 

SPIAT3D does not utilise any protein or mRNA markers, so you don't to worry
about that when converting to a SpatialExperiment object, at least, not with
SPIAT3D.

The SpatialExperiment function from the SpatialExperiment package allows you to
create a SpatialExperiment object and it has 3 necessary inputs:

1. assay: this would be your assay data. This might be from a Seurat object,
containing your protein marker information. But as stated before, SPIAT3D won't
use this information, so you can either include it, or set the assay parameter
to an empty assay, shown in the example below.

2. colData: this would be your data frame.

3. spatialCoordsNames: this would be the column names of your data frame which
correspond to the x, y, and z coordinates of each cell.

You can also include metadata - this might be extra information attached to your
original data that you still want to keep with the SpatialExperiment object.
This is shown in the example below.

```{r}
# Sample data frame with x,y,z coords and cell type identity
df <- data.frame("x" = runif(10),
                 "y" = runif(10),
                 "z" = runif(10),
                 "cell_type" = c(rep("Tumour", 5), rep("Immune", 5)))

head(df)

# Example metadata
df_metadata <- list("stuff", c(1, 2, 3))

# This code converts the df into a SpatialExperiment object spe
spe <- SpatialExperiment::SpatialExperiment(
  assay = matrix(data = NA, nrow = 0, ncol = nrow(df)), # ncol must match #cells in your data
  colData = df,
  spatialCoordsNames = c("x", "y", "z"),
  metadata = df_metadata)
```

# Accessing information from a SpatialExperiment object
SPIAT3D is able to obtain the x,y,z coordinates and cell type identity of each
cell in the SpatialExperiment object, to use for subsequent analysis. You can do
that to by yourself too.

You can extract the x,y,z coordinates and cell type identity of each cell. If
you put any metadata or an assay in the SpatialExperiment object, you can get
that back too.
```{r}
# Getting the coordinates
spe_coords <- SpatialExperiment::spatialCoords(spe)
head(spe_coords)

# Getting the cell type identity
# This depends on the original name of the column of the data frame containing 
# the cell types
cell_types <- spe$cell_type 
head(cell_types)

# Getting the metadata
spe_metadata <- spe@metadata
head(spe_metadata)

# Getting the assay
# This code might be wrong fyi, I'm not an expert of getting the assay
spe_assay <- spe@assays@data[[1]]
head(spe_assay)

```
# Updating a Spatial Experiment object so it is compatible with SPIAT3D
SPIAT3D requires your SpatialExperiment object is formatted in a specific way,
particularly in regards to the x,y,z coordinates.

The naming used to refer to the x,y,z coordinates must be "Cell.X.Position",
"Cell.Y.Position" and "Cell.Z.Position", for the x,y,z coordinates respectively.
It's named this way to mirror the naming in SPIAT. The code below shows you how
to do this.

```{r}
# Changing the names of the x,y,z coordinates so it works with SPIAT3D
colnames(SpatialExperiment::spatialCoords(spe)) <- c("Cell.X.Position", "Cell.Y.Position", "Cell.Z.Position")
head(SpatialExperiment::spatialCoords(spe))
```
# Note about SPIAT3D functions and feature_colname
All of the SPIAT3D functions have a parameter called 'feature_colname', which
refers to:

The original name of the column of the data frame containing the cell types. Or,
the name of the column in the SpatialExperiment object containing the cell 
types.

In our example, this is "cell_type". SPIAT3D default feature_colname to 
"Cell.Type". You just have to make sure to input the correct name that matches
with your SpatialExperiment object. You can also change your SpatialExperiment
object to use "Cell.Type" instead, as shown below.

```{r}
# Printing colData of SpatialExperiment object, it should include what you have
# used to refer to your cell types.
print(colnames(SummarizedExperiment::colData(spe)))

# Changing name of cell types to "Cell.Type", I will keep the other column name
# the same.
colnames(SummarizedExperiment::colData(spe)) <- c("Cell.Type", "sample_id")
print(colnames(SummarizedExperiment::colData(spe)))
```

