---
title: "Clustering algorithms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clustering algorithms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SPIAT3D)
```

# Introduction
SPIAT3D has three clustering algorithms to identify cell clusters within 3D
spatial data. Some have been adapted from the original SPIAT, some are new. 

This vignette will go through each of them. Firstly, you will need 3D
spatial data to analyse, I will be using a simulated dataset available in this
package.

```{r}
# Get simulated SpatialExperiment object to use as an example for analysis
simulated_spe <- readRDS(system.file("extdata", "simulated_spe.RDS", package = "SPIAT3D"))

print(simulated_spe)

```

# Alpha hull clustering
INSERT TEXT HERE

```{r}
# # Alpha hull clustering
# alpha_hull_spe <- alpha_hull_clustering3D(
#     spe = simulated_spe,
#     cell_types_of_interest = c("Tumour", "Immune"),
#     alpha = 8,
#     minimum_cells_in_cluster = 5,
#     feature_colname = "Cell.Type",
#     plot_image = TRUE
# )
# 
# # Can also plot alpha hull clusters separately with your own colours :) for
# # each cell type. No, you can't change the color of the alpha hulls :(
# fig <- plot_alpha_hull_clusters3D(
#     spe_with_alpha_hull = alpha_hull_spe,
#     plot_cell_types = c("Tumour", "Immune", "Endothelial", "Others"),
#     plot_colours = c("orange", "skyblue", "tomato", "lightgray"),
#     feature_colname = "Cell.Type"
# )
# methods::show(fig)
```


# DBSCAN clustering
INSERT TEXT HERE

```{r}
# DBSCAN clusters
dbscan_spe <- dbscan_clustering3D(
    spe = simulated_spe,
    cell_types_of_interest = c("Tumour", "Immune"),
    radius = 8,
    minimum_cells_in_radius = 10,
    minimum_cells_in_cluster = 30,
    feature_colname = "Cell.Type",
    plot_image = TRUE
)

# Plotting separately so it appears on GitHub pages, but the function should do
# this if you set plot_image = TRUE
dbscan_spe$dbscan_cluster <- ifelse(dbscan_spe$dbscan_cluster == 0, "non_cluster", paste("cluster_", dbscan_spe$dbscan_cluster, sep = ""))
plot_cells3D(dbscan_spe, feature_colname = "dbscan_cluster")
```


# Grid based clustering
INSERT TEXT HERE

```{r}
# # Grid based clusters
# grid_based_spe <- grid_based_clustering3D(
#     spe = simulated_spe,
#     cell_types_of_interest = c("Tumour", "Immune"),
#     n_splits = 10,
#     minimum_cells_in_cluster = 30,
#     feature_colname = "Cell.Type",
#     plot_image = TRUE
# )
# 
# # Can also plot grid based clusters separately with your own colours :) for each
# # cell type. No, you can't change the color of the prisms which form the 
# # clusters :(
# fig <- plot_grid_based_clusters3D(
#     spe_with_grid = grid_based_spe,
#     plot_cell_types = c("Tumour", "Immune", "Endothelial", "Others"),
#     plot_colours = c("orange", "skyblue", "tomato", "lightgray"),
#     feature_colname = "Cell.Type"
# )
# methods::show(fig)
```

# Further analysis of the clusters
Once you have identified the clusters, SPIAT3D offers several functions that can
help analyse certain aspects of the clusters. These include identifying the
cells that make up each cluster, identifying the cells that border each cluster,
determining how close other cells are to the cluster and calculating the
approximate centre and volume of each cluster.


