---
title: "Distance based metrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distance based metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SPIAT3D)
```

# Introduction
SPIAT3D has two distance based metrics that have been adapted from the original
SPIAT. This vignette will go through both of them. Firstly, you will need 3D
spatial data to analyse. I will be using a simulated dataset available in this
package.

```{r}
# Get simulated SpatialExperiment object to use as an example for analysis
simulated_spe <- readRDS(system.file("extdata", "simulated_spe.RDS", package = "SPIAT3D"))

print(simulated_spe)

plot_cells3D(simulated_spe,
             plot_cell_types = c("Tumour", "Immune", "Endothelial", "Others"),
             plot_colours = c("orange", "skyblue", "tomato", "lightgray"))
```

# Pairwise distance between cells
The first distance based metric is the 'pairwise distance between cells' metric
which involves computing and tabulating the all distances between all cell types
that the user is interested in. This outputs a distance matrix, showing the
distance for each cell-cell pair.

Be warned when using this function, its very storage and time demanding, and it
may not be suitable for use for large datasets.

```{r}
# Calculate pairwise distance between cells
pairwise_distances <- calculate_pairwise_distances_between_cell_types3D(
    spe = simulated_spe,
    cell_types_of_interest = c("Tumour", "Immune"),
    feature_colname = "Cell.Type",
    show_summary = FALSE,
    plot_image = FALSE
)
```

# Minimum distance between cells
The second distance based metric is the 'minimum distance between cells' metric
which involves computing and tabulating only the minimum distances between the
cell types which the user is interested in. This outputs a similar, but smaller
distance matrix compared to the pairwise distance between cells function, 
showing the minimum distance for each cell-cell pair.

```{r}
# Calculate minimum distance between cells
minimum_distances <- calculate_minimum_distances_between_cell_types3D(
    spe = simulated_spe,
    cell_types_of_interest = NULL,
    feature_colname = "Cell.Type",
    show_summary = FALSE,
    plot_image = FALSE
)
```

# Interpretating the distance matrices
After obtaining the distance matrices from either the pairwise or minimum
distance functions, SPIAT3D offers ways to interpret the data. This includes
printing a summary of the distance data, showing the mean, median, ... for each
cell-cell pair. This also includes plotting the distance data for each cell-cell
pair as violin plots.
```{r}
# Print summary of pairwise distance data
distances_summary <- summarise_distances_between_cell_types3D(
    distances_df = pairwise_distances
)
print(distances_summary)

# Plot violin plots of minimum distance data
fig <- plot_distances_between_cell_types_violin3D(
    distances_df = minimum_distances,
    scales = "free_x"
)
methods::show(fig)
```

# Further analysis of distance matrices
For our own spatial analysis, we found that it was a good idea to represent the
3D spatial data using a single value, rather than a large matrix of values. How
you choose to interpret the distance matrices is up to you.

What we adopted to use was the average minimum distance between cells (AMD). We
ignored pairwise distance as this was a bit too slow to run. Calculating AMD on
several 3D datasets gives you a quantitative way to compare the datasets.
```{r}
# Get Tumour/Immune AMD
AMD <- distances_summary$mean[distances_summary$pair == "Tumour/Immune"]
print(AMD)
```




